\section{Negative Results: Direct Delta Prediction}
\label{sec:negative_results}

The natural response to the Phase~8 geometry finding is to predict the
state-space delta directly:
\begin{equation}
    \Delta z_t = f_{\bar{\theta}}(s_{t+1})
               - f_{\bar{\theta}}(s_t).
    \label{eq:explicit_delta}
\end{equation}
This section records why that did not work.

\subsection{Initial Three-Way Comparison}

The initial 15K-step comparison tested standard JEPA prediction, an
early-layer delta readout, and a final-layer delta readout.  The baseline
won decisively:

\begin{table}[h]
\centering
\caption{Phase~8 direct delta prediction comparison.}
\label{tab:delta}
\smallskip
\begin{tabular}{llll}
\toprule
Metric & Baseline JEPA & Delta early-layer & Delta final-layer \\
\midrule
Validation delta cosine & $0.972$ & $0.278 \rightarrow 0.388$ & $0.241 \rightarrow 0.271$ \\
Validation cosine & $0.980$ & $0.106 \rightarrow -0.09$ & $-0.006 \rightarrow 0.04$ \\
Lift over copy & $+0.280$ & $-0.648 \rightarrow -0.84$ & $-0.910 \rightarrow -0.82$ \\
Train delta cosine & $0.985$ & $0.500$ & $0.499$ \\
\bottomrule
\end{tabular}
\end{table}

Early-layer readout was better than final-layer readout, confirming the
geometry probe, but neither was competitive with the JEPA baseline.

\subsection{Fair-Loss Fixes}

An implementation audit found a real asymmetry: the baseline optimized a
directional normalized objective, while the delta variants used a raw
LogCosh loss on tiny deltas.  Two fixes were tested:
\begin{enumerate}
    \item normalized delta loss;
    \item normalized delta loss plus a residual predictor.
\end{enumerate}
The residual variant improved validation delta cosine to about $0.356$
around step 2800, but it did not break through.  The overall conclusion
is therefore stronger than ``the first implementation was unfair.''
After correcting the loss asymmetry, direct delta prediction still
remained far below the online-to-target prediction baseline.

\subsection{Interpretation}

The negative result is not incidental.  The explicit delta lives in a
space where the signal is tiny, low-rank, and weakly correlated across
steps.  The successful model does not recover by making that delta target
cleaner.  It recovers by preventing the encoder from destroying useful
intermediate representations and by learning the translation between
online and target spaces.
